---
title: "Statistical Computing HW4"
author: "Wei-Chen Chang r12227118"
date: "Due: 2024-05-08"
output:
  pdf_document:
    fig_caption: yes
    extra_dependencies: amsmath
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width= 4.8,fig.height=3.2, fig.align = "center")

```

We want to sample from the double gamma distribution ($\alpha$ = 2 ) with pdf
$$
f(x|\alpha)=\frac{1}{2\Gamma(\alpha)}|x|^{\alpha-1}e^{-|x|};\;-\infty<x<\infty
$$
1. Use the acceptance-rejection algorithms with the proposal distribution
$Y \sim t(2)$. Plot the scaled histogram and matches it up with the theoretical PDF.  
2. Use the Metropolis-Hastings algorithm with the proposal distribution $Y \sim N(X_t, 10)$. Plot the scaled histogram and matches it up with the theoretical PDF.  
3. Compare the acceptance rate and performance of two algorithms.  


# Ans:

## Setup
```{r import_pack, message=FALSE}
library(tidyverse)
library(glue)
Nsim <- 5000
doub_gamma <- function(x, alpha=2){
  return(1/(2*gamma(alpha)) *abs(x)^(alpha-1)*exp(-abs(x)))
}
```

Set number of iteration = `r Nsim`, and define the function `doub_gamma()` of double gamma pdf.  
For visualization purpose, see Fig.\ref{fig:doub_gamma}.

```{r doub_gamma, fig.cap="Target and Proposed Distribution in AR Method"}
curve(doub_gamma, -10, 10, col="red", ylab= "density", ylim = c(0,.35),
      main = "Double Gamma and t Distribution")
curve(dt(x, df=2),-10, 10, lty=1, add =TRUE)
curve(3*dt(x, df=2),-10, 10, lty=2, add =TRUE, col = "grey20")
legend("topright",
       legend = c("doub_gamma",expression(t(df=2)),expression(3%*%t(df=2))),
       col = c("red", "black", "grey20"), lty = c(1,1,2), cex=.6,xjust = 0, bty="n")
```
## 1. A-R method

After trial and error, set c = 3 seems to be appropriate (see Fig.\ref{fig:doub_gamma}).
The result can be seen in Fig.\ref{fig:AR}.

```{r AR, results='asis', fig.cap= "AR Sampler and Theoretical pdf"}
AR_sim <- function(N = Nsim, c=3){
  env_samp <-  rt(N, df = 2) # envelop
  u <- runif(Nsim)
  prop <- doub_gamma(env_samp)/(c*dt(env_samp, df=2)) # acceptance crit
  samp <- env_samp[u<prop] #valid sample
  result <- list(sample = samp,
       accept_rate = length(samp)/Nsim)
  return(result)
}
plot_sim <- function(sample, title = "Sample",n=Nsim){
  sample |> hist(breaks = 100, freq=FALSE, xlab="x",
                  main = glue("Histogram of {title},(n={n})"))
  curve(doub_gamma, -10, 10, col="red", add= TRUE)
}
AR_samp <- AR_sim()
plot_sim(AR_samp$sample, "AR method")
```

## 2.  Metropolis-Hastings algorithm

Accept Criterion:
$$
\alpha = \min\{\frac{f(Y)q(Y|X_{i-1})}{f(X_{i-1})q(X_{i-1}|Y)} ,1\}
$$
where $q(x|y)$ is the pdf of $N(y, \sigma=10)$.

```{r, results='asis'}
MH_sim <- function(N=Nsim){
  MH_samp <- 1:Nsim
  MH_samp[1] <- rnorm(1,sd=10) # x0
  u <- runif(Nsim)
  accept <- 1 # accept first sample
  for (i in 2:Nsim){
    x <- MH_samp[i-1]
    y <- rnorm(1, mean = x, sd=10) 
    num <- doub_gamma(y) * dnorm(y, mean=x, sd=10)
    den <- doub_gamma(x) * dnorm(x, mean=y, sd=10)
    if (u[i] < (num/den)){
      MH_samp[i] <- y
      accept <- accept+1
    }else MH_samp[i] <- x # reject
  }
  result <- list(sample = MH_samp, accept_rate=accept/Nsim)
  return(result)
}
MH_samp <- MH_sim()
```

First check its convergence through caterpillar plot. As the left panel of Fig.\ref{fig:caterp} shows,
the convergence seems to be OK.
The histogram of the M-H sampler can be seen in the right panel of Fig.\ref{fig:caterp}.
```{r caterp, fig.cap="M-H sampler",fig.width=7.2}
#Check for Convergence
par(mfrow=c(1,2))
MH_samp$sample |> plot(type = 'l', main=glue("n={Nsim}, burnin=0"),
                       ylab=expression(x[i]), xlab="i")
plot_sim(MH_samp$sample, title = "M-H method")
```

One can see M-H performs fine when we set number of iteration $(i)$ as `r Nsim`,
even no burn-in sample were discard.

## 3. Comparison b/t 2 mehtods

Acceptance rate in 2 methods are shown below. One can notice M-H method has lower
acceptance rate compared to AR method.
```{r, results='asis'}
cat(glue("Acceptance rate in AR method: {AR_samp$accept_rate*100}%\n
         Acceptance rate in M-H method: {MH_samp$accept_rate*100}%"))
```

For performance evaluation, repeating the sampling procedure for 100 times.

```{r comp, results='hold'}
cat(glue("Time taken in AR method (N = {Nsim}) for 100 times: \n
         "))
system.time(for (i in 1:100) AR_sim())
cat(glue("Time taken in M-H method (N = {Nsim}) for 100 times: \n
         "))
system.time(for (i in 1:100) MH_sim())
```
One can see M-H is about 10 times slower than AR method in this case.
