---
title: "Statistical Computing HW5"
author: "Wei-Chen Chang r12227118"
date: "Due: 2024-05-22"
output:
  pdf_document:
    fig_caption: yes
    extra_dependencies: amsmath
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width= 5.6,fig.height=4.0, fig.align = "center")
```

```{r jags, eval=FALSE, include=FALSE}
for (p in c("R2jags", "coda","rjags")){
  install.packages(p,dependencies = TRUE)
}
```

# Setup

## The Model

$$
E[Y_i]=\frac{1}{e_i} = \beta_0 + \frac{\beta_1}{(N_i+\alpha_1)}+
\frac{\beta_2}{(P_i+\alpha_2)}+\frac{\beta_3}{(K_i+\alpha_3)}
$$
The nutrient values of different levels in  $N,P, K$ are N=0, 100, 200, 400,
P=0, 22, 44, 88, and K=0, 42, 84, 168. respectively.
168.

And the list of priors:
$$
\begin{cases}
Y_i \sim gamma(\nu,\nu e_i)\\
\nu\sim gamma(0.01, 0.01)\\
\alpha_1 \sim N(\mu=40, \sigma^2 = 100)\\ 
\alpha_2 \sim N(\mu=22, \sigma^2 = 100)\\ 
\alpha_3 \sim N(\mu=32, \sigma^2 = 100)\\ 
\beta_0 \sim N(\mu=0, \sigma^2 = 10000)\\
\beta_j \sim N(\mu=0, \sigma^2 = 10000)I[\beta_j>0],\; j=1,2,3 \\
\end{cases}
$$


### Specifying Model in JAGS
```{r}
# Load Dataset
data(bermuda.grass, package = "LearnBayes")
b.grass.list <- c(as.list(bermuda.grass), 
                  list("N" = nrow(bermuda.grass))
)
# JAGS model
modelstring ="
model{
  for(i in 1:N){
    y[i] ~ dgamma(nu, mu[i]); mu[i] <- nu*eta[i]
    eta[i] <- 1/yhat[i] 
    yhat[i] <- beta0 + (beta1/(100*Nit[i]+alpha[1]))+(beta2/(22*Phos[i]+alpha[2]))+(beta3/(42*Pot[i]+alpha[3]))
  }
## priors ##
  nu~dgamma(0.01, 0.01)
  alpha[1]~dnorm(40, 1/100)
  alpha[2]~dnorm(22, 1/100)
  alpha[3]~dnorm(32, 1/100)
  beta0~dnorm(0, 1/10000)
  beta1~dnorm(0, 1/10000) T(0,) # truncated below 0
  beta2~dnorm(0, 1/10000) T(0,) # truncated below 0
  beta3~dnorm(0, 1/10000) T(0,) # truncated below 0
  
}
"
writeLines(modelstring, con="bermuda.bug")
```

# Analysis

Here `R2jags` package is used to done the MCMC. The Posterior Mean, S.D., and other statistics
can be seen in the output. One can see that every parameters have `Rhat` ($\hat R$, Gelman-Rubin Statistic) 
values lower than 1.01, the convergence of the MCMC seems to be OK. 
```{r message=FALSE}
library(coda)
library(rjags)
library(R2jags)
param.names <- c("nu", "alpha", paste0("beta",0:3))
# set initial values
bayes.mod.inits <- function(){
 list("nu" = rgamma(1, 0.1, 0.1),
      "alpha" = c(rnorm(1, 40, sd=5), 
                  rnorm(1, 22, sd=5),
                  rnorm(1, 32, sd=5)),
      "beta0" = rnorm(1, sd=5), # such initial for beta0 seems to generate error
      "beta1" = abs(rnorm(1,0, sd=5)),
      "beta2" = abs(rnorm(1,0, sd=5)),
      "beta3" = abs(rnorm(1,0, sd=5))
      )
} 
inits <- list(bayes.mod.inits(), bayes.mod.inits(), bayes.mod.inits())
set.seed(907)
ber.jags<- jags(data = b.grass.list,
     # inits = inits, # Note: errors happends a lot when using self-specified inits
     parameters.to.save = param.names,
     n.chains=3, n.iter=10000, n.burnin=1000,
     model.file = "bermuda.bug")
ber.jags |> print()
```
For visulaizing purpose, the posterior density plot were shown below.
```{r}
library(lattice)
ber.mcmc <- as.mcmc(ber.jags)
densityplot(ber.mcmc, layout=c(3,3), aspect="fill", 
       main = "Density Plot")
```
One can see the density plot of 3 chains seems aligned.

## Model Diagnosis

### Convergence
For diagnosis of the Markov Chains, here Geweke statistic were computed and
convergence plot were drawn to check for convergence.  

```{r}
geweke.diag(ber.mcmc)
```
Geweke statistic larger than 2.5 or smaller than -2.5 is a warning sign for nonstationarity   
samples. One can see only `beta2` from 2nd chain suffers such.

```{r}
xyplot(ber.mcmc, layout=c(3,3), aspect="fill", 
       main = "Divergence Plot")
```
The Divergence plot also indicate that the chains were OK for convergence.

### ACF
And Autocorrelation plot were plotted below. One can see the autocorrelation after 5 lags are
mostly between -.1 to .1. It support the samples were uncorrelated. 
```{r ACF, fig.width= 6.0,fig.height=4.8}
acfplot(ber.mcmc, main = "ACF Plot", layout=c(3,3), aspect="fill")
```





```{r eval=FALSE, include=FALSE}
library(coda)
library(rjags)

ber.mod <- jags.model(file="bermuda.bug",
                      inits = inits,
                      data = b.grass.list, n.chains=3 , n.adapt=500)
update(ber.mod, n.iter=1000, progress.bar="text") 
out=coda.samples(ber.mod,
                 variable.names=param.names,
                 n.iter=10000, 
                 thin=5) 
summary(out)
``` 
