---
title: "Statistical Computing HW3"
author: "Wei-Chen Chang r12227118"
date: "Due: 2024-05-01"
output:
  pdf_document:
    fig_caption: yes
    extra_dependencies: amsmath
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width= 5,fig.height=3.5, fig.align = "center")
library(tidyverse)
library(glue)
```

Consider the `scor (bootstrap)` test score data on 88 students who took
examinations in five subjects. The five-dimensional scores data have a 5 × 5
covariance matrix $\Sigma$, with positive eigenvalues $\lambda_1>...,>\lambda_5$.
In principal components analysis,

$$
\theta = \frac{\lambda_1}{\Sigma_{i=1}^{5}\lambda_i}
$$

measures the proportion of variance explained by the first principal component.
Let $\hat{\lambda}_1>...,>\hat\lambda_5$ be the eigenvalues of $\Sigma$, where $\hat\Sigma$ is the MLE of $\Sigma$.  

Compute the sample estimate
$$
\hat\theta = \frac{\hat\lambda_1}{\Sigma_{i=1}^{5}\hat\lambda_i}
$$

of $\theta$.

1. Use bootstrap to estimate the bias and standard error of $\hat\theta$.  

2. Obtain the jackknife estimates of bias and standard error of $\hat\theta$.  

3. Compute 95% percentile and BCa confidence intervals for $\hat\theta$.  

# Ans:

First, the MLE of $\Sigma$, $\hat\Sigma$ is computed as:
$$
\hat\Sigma = \frac{1}{n}\mathbf{(x - \bar{x})^T(x - \bar{x})},
$$
where $\mathbf{x}$ is a $88\times5$ matrix , and $\mathbf{\bar x} = \frac{1}{N}\mathbf{11^Tx}$,
where $\mathbf{1}_{88\times 1}= \begin{pmatrix}1 &1&...&1\end{pmatrix}^T$.

Here Compute MLE first and the sample estimate first.
```{r, results='asis'}
data(scor, package = "bootstrap")
dta <- as.matrix(scor)
N <- nrow(scor)

get_fPC <- function(mtx){
  N <- nrow(mtx)
  mlecov <-  cov(mtx)*(N-1)/N # give MLE
  eig_val <- eigen(mlecov)$value
  first_PC <- eig_val[1]/sum(eig_val)
  return(first_PC)
}
theta <- get_fPC(scor) #theta hat
cat(glue("$\\hat\\theta={theta}$"))
```

## 1. bootstraping 

```{r set Nboot, include=FALSE}
Nboot <- 2000
```

Construct bootstrap samples $X^{*b}$, $b=1,..,B,$ $B=`r Nboot`$; $\hat\theta^{*b}=T(X^{*b})$ is the estimate computed from bootstrap sample$X^{*b}$.  
Bootstrap bias and s.e. can be computed as :

$$
\begin{aligned}
\widehat{bias}_{boot}(\hat\theta) &= \overline{\hat{\theta}^*}-\hat{\theta},\\
\widehat{se}_{boot}(\hat\theta) &= \sqrt{\frac{1}{B-1}\Sigma_{b=1}^{B}(\hat{\theta}^{*b} - \overline{\hat{\theta}^*})^2}\\
\end{aligned}
$$

```{r boot,results='asis'}
set.seed(989)
Nboot <- 2000
boot_theta <- 1:Nboot
for (i in 1:Nboot){
  boot <- dta[sample(1:N, N, replace = T),] #samp w/ replacement
  boot_theta[i] <- get_fPC(boot)
}
# hist(boot_theta, breaks = 100,freq=F);abline(v = theta, col="red",lwd=2, lty=2)
b.bias <- mean(boot_theta)-theta
b.se <- sd(boot_theta)
cat(glue("Bootstrap Bias (B={Nboot}) = {b.bias}, \n 
         Bootstrap s.e. (B={Nboot}) = {b.se}"))

```

## 2. Jackknife

Jackknife bias and s.e. are defined and coumputed as below: 
$$
\begin{aligned}
\widehat{bias}_{jack} &= (n-1)(\overline{\hat\theta_{(\cdot)}} - \hat\theta),\\ 
\widehat{se}_{jack} &= \sqrt{\frac{n-1}{n}\Sigma_{i=1}^{n}(\hat\theta_{(i)} - \overline{\hat\theta_{(\cdot)}})^2} 
\end{aligned}
$$

```{r, results='asis'}
jack_theta <- 1:N
for (i in 1:N){
  jack <- dta[-i,] #LLO
  jack_theta[i] <- get_fPC(jack)
}
j.bias <- (N-1)*(mean(jack_theta)-theta)
j.se <- sqrt(((N-1)/N)*sum((jack_theta- mean(jack_theta))^2))

cat(glue("Jackknife Bias ={j.bias}, \n
         Jackknife s.e.={j.se}"))
```

## 3. CI

Compute 95% percentile and BCa confidence intervals
```{r, results='asis'}
conf_level <- .05 # Type 1 error rate
alpha <- c(conf_level/2, 1- (conf_level/2))
zalpha <- qnorm(alpha)
# Percentile 95% CI
p.CI <- quantile(boot_theta, alpha, type=1) 
# BCa
## bias correction:
bias <- qnorm(mean(boot_theta - theta>0))
## accelaration
delt.jack <- mean(jack_theta)-jack_theta
accer <- sum(delt.jack^3)/(6*(sum(delt.jack^2))^(3/2))
bca.perc <- (bias+ ((bias+zalpha)/(1-accer*(bias+zalpha))))%>% 
  pnorm()
#BCa 95% CI
bca.CI <- quantile(boot_theta, bca.perc, type=6) 

cat(glue("95% percentile CI: ({p.CI[1]},
               {p.CI[2]}),\n
               95% BCa CI: ({bca.CI[1]},
               {bca.CI[2]})"))
```

## `boot` package:

Additionally, here utilized `boot` to attain bootstrap bias, SE and `boot.ci` for Compute 95% percentile and BCa confidence intervals.
```{r, results='asis', warning=FALSE}
get_fPC2 <- function(mtx,i){
  N <- nrow(mtx)
  mlecov <-  cov(mtx[i, ])*(N-1)/N # give MLE
  eig_val <- eigen(mlecov)$value
  first_PC <- eig_val[1]/sum(eig_val)
  return(first_PC)
}
# package
library(boot)
set.seed(989)
boot2 <- boot(dta,get_fPC2, R = Nboot)
cat(glue("Bias: {mean(boot2$t) -boot2$t0},\n
         s.e.: {sd(boot2$t)}"))


ci_result <- boot.ci(boot2, type = c("base", "perc", "bca"))

cat(glue("95% percentile CI: ({ci_result$percent[4]},
               {ci_result$percent[5]}),\n
               95% BCa CI: ({ci_result$bca[4]},
               {ci_result$bca[5]})"))
```


### Visualization

The Histogram below visualized the distribution of Bootstrap Estimates, its 95% percentile CI,
95% BCa CI and the sample estimate $\hat\theta$.

```{r}
hist(boot2$t, breaks = 100, main = glue("Bootstrap Estimates (B={Nboot})"),
     xlab= expression(widehat(theta)^b)) 
abline(v = boot2$t0, col = "red", lwd=2)
abline(v = ci_result$percent[c(4,5)], col = "darkorange", lwd=2, lty=2)
abline(v = ci_result$bca[c(4,5)], col = "darkcyan", lwd=2, lty=3)
legend("topleft",
       legend = c(expression(widehat(theta)), "Percentile CI", "BCa CI"),
       col = c("red", "darkorange", "darkcyan"), lty = 1:3, lwd=2, cex=.5)
```


